---
title: "Week 02 - More dplyr"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

## Exercise 01: summarise

Use [sleep](https://ggplot2.tidyverse.org/reference/msleep.html) dataset from the `ggplot2` package.

Compute the mean total sleep time across all animals. Use `summarise`.

```{r exercise 01-1}

```

Compute the mean and standard deviation of `sleep_total` in a single summarise call.

```{r exercise 01-2}

```

## Exercise 02: group_by + summarise

Group the dataset by `vore.` For each group, calculate the mean and median of `sleep_total`. Use `group_by`.

```{r exercise 02-1}

```

Group the dataset by conservation status. For each group, compute the number of species and their mean body weight.

```{r exercise 02-2}

```

## Exercise 03: arrange + group_by

Which animal(s) in each `vore` category sleep the most? Hint: group by `vore`, then arrange by `sleep_total` descending, then use `slice_max`.

```{r exercise 03-1}

```

## Exercise 04: mutate + case_when

Create a new variable called `sleep_level`, with three levels: `short_sleep` if `sleep_total` is less than 8 hours, `average_sleep` if `sleep_total` is between 8 and 12, and `long_sleep` if they sleep over 12 hours.

```{r exercise 04-1}

```

## Exercise 05: across()

Select only the numeric columns and compute their column means (ignore NAs).

```{r exercise 05-1}

```

## Exercise 06: chaining all together

Now filter only mammals heavier than 10 kg, group them by `vore`, compute their mean sleep time, and arrange the results by sleep in descending order.

```{r exercise 06-1}

```
