---
title: "Week 02 - More dplyr - SOLUTION"
author: "Karim Rivera-Lares"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

## Exercise 01: summarise

Use [sleep](https://ggplot2.tidyverse.org/reference/msleep.html) dataset from the `ggplot2` package.

Compute the mean total sleep time across all animals. Use `summarise`.

```{r exercise 01-1}
data("msleep")
msleep %>% 
  summarise(mean_sleep = mean(sleep_total))
```

Compute the mean and standard deviation of `sleep_total` in a single summarise call.

```{r exercise 01-2}
msleep %>% 
  summarise(mean_sleep = mean(sleep_total, na.rm = TRUE),
            sd_sleep = sd(sleep_total, na.rm = TRUE))
```

## Exercise 02: group_by + summarise

Group the dataset by `vore`, For each group, calculate the mean and median of `sleep_total`. Use `group_by`.

```{r exercise 02-1}
msleep %>% 
  group_by(vore) %>% 
  summarise(mean_sleep = mean(sleep_total),
            median_sleep = median(sleep_total))
```

Group the dataset by conservation status. For each group, compute the number of species and their mean body weight.

```{r exercise 02-2}
msleep %>% 
  group_by(conservation) %>% 
  summarise(n_species = n(),
            mean_weight = mean(bodywt, na.rm = TRUE))
```

## Exercise 03: arrange + group_by

Which animal(s) in each `vore` category sleep the most? Hint: group by `vore`, then arrange by `sleep_total` descending, then use `slice_max`.

```{r exercise 03-1}
msleep %>% 
  group_by(vore) %>% 
  slice_max(order_by = sleep_total, n = 1)
```

## Exercise 04: mutate + case_when

Create a new variable called `sleep_level`, with three levels: `short_sleep` if `sleep_total` is less than 8 hours, `average_sleep` if `sleep_total` is between 8 and 12, and `long_sleep` if they sleep over 12 hours.

```{r exercise 04-1}
msleep %>% 
  mutate(sleep_level = case_when(
    sleep_total < 8 ~ "short_sleep",
    sleep_total >= 8 & sleep_total <= 12 ~ "average_sleep",
    sleep_total > 12 ~ "long_sleep"
  ))
```

## Exercise 05: across()

Select only the numeric columns and compute their column means (ignore NAs).

```{r exercise 05-1}
msleep %>% 
  summarise(across(where(is.numeric), ~mean(.x, na.rm = TRUE)))
```

## Exercise 06: chaining all together

Now filter only mammals heavier than 10 kg, group them by `vore`, compute their mean sleep time, and arrange the results by sleep in descending order.

```{r exercise 06-1}
msleep %>%
  group_by(vore) %>%
  summarise(mean_sleep = mean(sleep_total, na.rm = TRUE)) %>%
  arrange(desc(mean_sleep))
```
