---
title: "Week 2: Whole game exercises SOLUTION"
author: "Karim Rivera-Lares"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

The following exercises are a quick overview of the main tools of data science: import, tidy, transform, visualise, model, and communicate data.

We will use the built-in `starwars` dataset from the `dplyr` package. Remember that `dplyr` is contained within `tidyvsere`, so once you have loaded `tidyverse` there is no need to load `dplyr`.

## Exercise 01: Import

Load the dataset and take a first look.

```{r exercise 01-1}
starwars
glimpse(starwars)
```

What do you notice? What type of variables can you see? How many observations and variables are there?

## Exercise 02: Tidy

Check if the dataset is tidy. There are three rules which make a dataset tidy: a) each variable must have its own column, b) each observation must have its own row, and c) each value must have its own cell. Take a look at the dataset and modify what you need to modify for there rules to be true. HINT: You can use `unnest()` to expand the columns that contain a list of values, but you can only unnest one variable at a time.

Look at whether there are missing values in each column. HINT: you can use summarise() and across().

```{r exercise 02-1}
starwars %>%
  summarise(across(everything(), ~sum(is.na(.))))

starwars_films <- starwars %>%
  unnest(films)
```

## Exercise 03: Tidy

Do some transformation in your data. For example, look for the BMI formula and create a variable with the characters' BMI, and create a new variable for height in meters.

```{r exercise 03-1}
starwars_bmi <- starwars %>%
  mutate(height_m = height / 100,
         BMI = mass / (height_m^2))

head(starwars_bmi)
```

## Exercise 04: Visualise

Create a scatterplot of height vs mass, and a boxplot of mass by gender.

```{r exercise 04-1}
ggplot(starwars, aes(x = height, y = mass)) +
  geom_point()

ggplot(starwars, aes(x = gender, y = mass)) +
  geom_boxplot()
```

## Exercise 05: Model

Fit a linear regression predicting mass from height.

```{r exercise 05-1}
lm_fit <- lm(mass ~ height, data = starwars)
summary(lm_fit)
```

## Exercise 06: Communicate

Summarise your findings: What are the main patterns you observed? How good is your model? Present your results in a short paragraph.

**Answer:** The linear model shows a verwy weak relationship between height and mass. The estimated slope is small (0.624), and the p-value (0.323) indicates it is not statistically significant. The R-squared is extremely low (0.017), meaning that height explains only about 1.7% of the variation in mass. Residuals are large, with one extreme positive residual, suggesting that the model does not fit the data well. Overall, while taller characters tend to weigh slightly more on average, height is a poor predictor of mass in this dataset, and the model should not be relied on for accurate predictions.
