---
title: "Solutions to practice exercises"
format: html
---

-   Define a function called **add_two_numbers()** that takes two arguments, adds them together, and returns the result.

    ```{r}
    add_two_numbers <- function(num_1, num_2) {
      num_1 + num_2
    }
     
    sum_3_4 <- add_two_numbers(3, 4)
    sum_3_4
     
     
    sum_5_5 <- add_two_numbers(5, 5)
    sum_5_5
    ```

-   Define a function called **square_number()** that takes a single argument and returns its square.

    ```{r}
    square_number <- function(num) {
      num**2
    }
     
    square_3 <- square_number(3)
    square_3
     
     
    square_5 <- square_number(5)
    square_5
    ```

-   Define a function called **is_even()** that takes one argument, checks if the number is even or odd, and returns TRUE for even numbers and FALSE for odd numbers. HINT use the modulo operator %%.

    ```{r}
    is_even <- function(x) {
      x %% 2 == 0
    }
     
    even_num_3 <- is_even(3)
    even_num_3
     
    even_num_4 <- is_even(4)
    even_num_4
    ```
