---
title: "Week 2 Tutorial Materials"
format: html
---

Here you will find the tutorial materials you will need for Week 2. To download the files, right click the link and select "Save link as...". Then save the .Rmd file in your folder.

## Tutorial materials for Thursday's lab

Student led discussion on Ethical Dilemmas in Health Care Data

**Exercise: Reading in Files**

```{r}
#| echo: false
#| warning: false
#| message: false
library(downloadthis)

download_link(
  link = "../documents/week-2/week2_set1_files_and_folders_exercise.Rmd",
  button_label = "Download the files and folders exercise",
  button_type = "danger",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)
```

**Exercise: warm-up.**

```{r}
#| echo: false
#| warning: false
#| message: false
library(downloadthis)

download_link(
  link = "../documents/week-2/week2_set3_warmup.Rmd",
  button_label = "Download the warm-up exercise",
  button_type = "danger",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)
```

## Tutorial materials for Friday's lab

**Exercise: dplyr**

```{r}
#| echo: false
#| warning: false
#| message: false
library(downloadthis)

download_link(
  link = "../documents/week-2/week2_set2_dplyr.Rmd",
  button_label = "Download the dplyr exercise",
  button_type = "danger",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)
```

**Exercise: More dplyr**

```{r}
#| echo: false
#| warning: false
#| message: false
library(downloadthis)

download_link(
  link = "../documents/week-2/week2_set2_more_dplyr.Rmd",
  button_label = "Download the second dplyr exercise",
  button_type = "danger",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)
```

**Exercise: Full Data Analysis Work Flow**

```{r}
#| echo: false
#| warning: false
#| message: false
library(downloadthis)

download_link(
  link = "../documents/week-2/week2_set5_whole_game_exercise.Rmd",
  button_label = "Download the second dplyr exercise",
  button_type = "danger",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)
```
