---
title: "Week 2: Whole game exercises"
author: "Karim Rivera-Lares"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

The following exercises are a quick overview of the main tools of data science: import, tidy, transform, visualise, model, and communicate data.

We will use the built-in `starwars` dataset from the `dplyr` package. Remember that `dplyr` is contained within `tidyvsere`, so once you have loaded `tidyverse` there is no need to load `dplyr`.

## Exercise 01: Import

Load the dataset and take a first look.

```{r exercise 01-1}

```

What do you notice? What type of variables can you see? How many observations and variables are there?

## Exercise 02: Tidy

Check if the dataset is tidy. There are three rules which make a dataset tidy: a) each variable must have its own column, b) each observation must have its own row, and c) each value must have its own cell. Take a look at the dataset and modify what you need to modify for there rules to be true. HINT: You can use `unnest()` to expand the columns that contain a list of values, but you can only unnest one variable at a time.

Look at whether there are missing values in each column. HINT: you can use summarise() and across().

```{r exercise 02-1}

```

## Exercise 03: Tidy

Do some transformation in your data. For example, look for the BMI formula and create a variable with the characters' BMI, and create a new variable for height in meters.

```{r exercise 03-1}

```

## Exercise 04: Visualise

Create a scatterplot of height vs mass, and a boxplot of mass by gender.

```{r exercise 04-1}

```

## Exercise 05: Model

Fit a linear regression predicting mass from height.

```{r exercise 05-1}

```

## Exercise 06: Communicate

Summarise your findings: What are the main patterns you observed? How good is your model? Present your results in a short paragraph.

**Answer:**
