---
title: "Week 1 Tutorial Materials"
format: html
---

During this course, it's recommended that you create a directory somewhere on your computer to store all the tutorial materials. You can call it whatever you like, but for the sake of simplicity, we recommend you create the directory on your Desktop and call it `data_science` (no spaces in the name).

## Thursday

Create a `New project` called `brfss_analysis` in your `data_science` directory. Save the starter file provided below in `brfss_analysis`.

Remember, you can use the R documentation to help you solve the exercises or Google it!

The goals of this exercise are:

-   Practice pair programming
-   Solve some problems using your current R knowledge (no Tidyverse!)

To download the files, right click the link and select "Save link as...". Then save the .Rmd file in your folder.

### BRFSS Analysis (Behavioral Risk Factor Surveillance System)

```{r}
#| echo: false
#| warning: false
#| message: false
library(downloadthis)

download_link(
  link = "../documents/week-1/1_brfss_analysis.Rmd",
  button_label = "Download the starter file",
  button_type = "danger",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)
```

## Friday

The goals of this exercise are:

-   Practice RMarkdown!

### RMarkdown Exercise

```{r}
#| echo: false
#| warning: false
#| message: false
library(downloadthis)

download_link(
  link = "../documents/week-1/2_markdown_exercise.Rmd",
  button_label = "Download the starter file",
  button_type = "danger",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)
```
